---
title: "ETC3250/5250 Assignment 1"
author: "Arindom Baruah (32779267)"
date: "2024-03-22"
quarto-required: ">=1.3.0"
format:
    html:
        output-file: assign01-submission.html
        css: "assignment.css"
execute: 
  echo: false
  message: false
  warning: false
---

```{r}

draw_confusion_matrix <- function(cm) {

  layout(matrix(c(1,1,2)))
  par(mar=c(2,2,2,2))
  plot(c(100, 345), c(300, 450), type = "n", xlab="", ylab="", xaxt='n', yaxt='n')
  title('CONFUSION MATRIX', cex.main=2)


  rect(150, 430, 240, 370, col='#3F97D0')
  text(195, 435, 'Bilby', cex=1.2)
  rect(250, 430, 340, 370, col='#F7AD50')
  text(295, 435, 'Quokka', cex=1.2)
  text(125, 370, 'Predicted', cex=1.3, srt=90, font=2)
  text(245, 450, 'Actual', cex=1.3, font=2)
  rect(150, 305, 240, 365, col='#F7AD50')
  rect(250, 305, 340, 365, col='#3F97D0')
  text(140, 400, 'Bilby', cex=1.2, srt=90)
  text(140, 335, 'Quokka', cex=1.2, srt=90)


  res <- as.numeric(cm$table)
  text(195, 400, res[1], cex=1.6, font=2, col='white')
  text(195, 335, res[2], cex=1.6, font=2, col='white')
  text(295, 400, res[3], cex=1.6, font=2, col='white')
  text(295, 335, res[4], cex=1.6, font=2, col='white')


  plot(c(100, 0), c(100, 0), type = "n", xlab="", ylab="", main = "DETAILS", xaxt='n', yaxt='n')
  text(10, 85, names(cm$byClass[1]), cex=1.2, font=2)
  text(10, 70, round(as.numeric(cm$byClass[1]), 3), cex=1.2)
  text(30, 85, names(cm$byClass[2]), cex=1.2, font=2)
  text(30, 70, round(as.numeric(cm$byClass[2]), 3), cex=1.2)
  text(50, 85, names(cm$byClass[5]), cex=1.2, font=2)
  text(50, 70, round(as.numeric(cm$byClass[5]), 3), cex=1.2)
  text(70, 85, names(cm$byClass[6]), cex=1.2, font=2)
  text(70, 70, round(as.numeric(cm$byClass[6]), 3), cex=1.2)
  text(90, 85, names(cm$byClass[8]), cex=1.2, font=2)
  text(90, 70, round(as.numeric(cm$byClass[8]), 3), cex=1.2)


  text(30, 35, names(cm$overall[1]), cex=1.5, font=2)
  text(30, 20, round(as.numeric(cm$overall[1]), 3), cex=1.4)
  text(70, 35, names(cm$byClass[11]), cex=1.5, font=2)
  text(70, 20, round(as.numeric(cm$byClass[11]), 3), cex=1.4)
}  
```



```{r}
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(caret)
```


## Exercises

#### 1. Basic math and computing (5pts)


```{r echo=TRUE}

S <- rbind(c(3,1),c(1,2))
X <- rbind(4,2)
A <- rbind(1,-1)


Matrix_calculation <- t(X-A) %*% solve(S) %*% (X-A)

```
The matrix calculation for computing $(X-A)^T S^{-1}(X-A)$ yields the result __`r Matrix_calculation[1,1]`__.


#### 2. ML concepts (8pts)

```{r}
d_pred <- read_csv("https://raw.githubusercontent.com/numbats/iml/master/data/pred_data.csv")
d_pred |> slice_head(n=3)
```



```{r}
model_1 <- d_pred %>% dplyr::select(c("y","pred1","bilby1","quokka1"))

model_1$y <- as.factor(model_1$y)
model_1$pred1 <- as.factor(model_1$pred1)
model_2 <- d_pred %>% dplyr::select(c("y","pred2","bilby2","quokka2"))
```

```{r}
cm_1 <- model_1 %>% count(y,pred1) %>% group_by(y) %>%
  mutate(cl_acc = n[pred1==y]/sum(n)) %>% pivot_wider(names_from = pred1,
                                                      values_from = n) %>%
  dplyr::select(y,bilby,quokka,cl_acc)
```

The accuracy for `model 1` which is calculated based on the number of true positives and the true negatives is observed to be __`r accuracy(model_1,y,pred1) %>% pull(.estimate) %>% round(3)`__.

However, the accuracy parameter for a model may not always be the best indicator. This is especially true when the data may contain unbalanced class distribution. In this case, we will rely on balanced accuracy which is based on the true positive and the true negative rate of prediction for the model.

The balanced accuracy for `model 1` is found to be __`r bal_accuracy(model_1,y,pred1) %>% pull(.estimate) %>% round(3)`__.

@fig-confmat1 illustrates the detailed confusion matrix for `model 1` with the 

```{r}
#| label: fig-confmat1
#| fig-cap: "Confusion matrix with key performance metrics of model 1"
model_1_cm <- confusionMatrix(model_1$y,model_1$pred1)

draw_confusion_matrix(model_1_cm)
```







#### 3. Visualisation (8pts)

YOUR ANSWER HERE

#### 4. Dimension reduction (15pts)

YOUR ANSWER HERE

## References

Hadley Wickham, Dianne Cook, Heike Hofmann, Andreas Buja
  (2011). tourr: An R Package for Exploring Multivariate
  Data with Projections. Journal of Statistical Software,
  40(2), 1-18. URL http://www.jstatsoft.org/v40/i02/.
  
Kuhn et al., (2020). Tidymodels: a collection of
  packages for modeling and machine learning using
  tidyverse principles. https://www.tidymodels.org  
  
OpenAI (2023). ChatGPT (version 3.5) [Large language model]. https://chat.openai.com/chat, full script of conversation [here]( https://chat.openai.com/share/746c4f77-9b7b-4c3b-867a-826f639b91aa)